Epidemiological birth-death models with partner notification

### 2024-09-05 article/article_plos.tex
minor

Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}). The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in a maximum-likelihood framework). In particular, it permits the estimation of the following epidemiological parameters:

@@

We propose an %non-Markovian 
extension of the MTBD models that adds partner notification (PN).  
In the partner notification process, an identified infected index case notifies their potentially infected contacts (e.g., sexual or drug injection partners for HIV), either on their own (patient referral) or with a help of a health service provider (provider referral, when a health adviser notifies the partners without identifying the index patient; or contract referral, when if the index case fails to notify partners within a specified time period, the health adviser proceeds to provider referral)~\cite{wardPartnerNotification2014}. The notified partners are then provided with appropriate tests and treatment. Therefore, a successful detection of a new case B via partner notification by an index case A implies several conditions to be met: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (ii) B is successfully reached and notified; (iii) B agrees to take a test; (iv) B is indeed infected. 
For instance, according to the report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013 for HIV, reaching the stage (ii) had an estimated probability of about 50\%; reaching the stage (iii) -- of about 41\% (as 83\% of all the patients who attended a clinic through PN had a reported HIV test); and the stage (iv) -- of less than 3\%.

@@

BD-PN hence has 3 classical BD and 2 additional parameters:
\begin{itemize}
 \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual with this rate. This leads to an internal node in the transmission tree;
 \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This leads to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability. This leads to a tip in the transmission tree being observed;
 \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability. This leads to the removal rate change for the partner branch in the transmission tree;
 \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via the standard procedure) gets removed \textit{and observed} with this rate. This leads to an observed tip in the transmission tree. 
\end{itemize}

@@

We applied bdpn to each of the trees fixing the sampling probability $\rho$ to its true value (for identifiability). %The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material. 
%As expected with a maximum-likelihood estimator, 
The tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.

@@

The PN test detected partner notification in this tree (p-value $=0.014$). Our bdpn estimator inferred the partner notification probability $\upsilon=0.09$ CI $[0.008-0.226]$,  
$R_e = 1.43$ $[1.26-1.66]$, infectious time $\frac{1}{\psi} = 8$ $[6-11]$ years, and 4.9 months [6 days -- 1.4 years] between partner notification and their sampling. These results globally agree (CIs intersect) with the $R_e$ estimate for this data using a coalescent model from Rasmussen \textit{et al.}~\cite{Rasmussen2017} (between 1.0 and 2.5), and the estimates using the BDSS model from Voznica~\textit{et al.}~\cite{Voznica2021}: $R_e$ of 1.6 and 1.7 and an infectious period of 10.2 and 9.8 years (depending on the tree representation used for the deep learner).

@@

For comparison, parameter estimation on the same tree with the BD model gave a slightly lower estimate of $R_e$ and a slightly shorter infectious time, though with intersecting CIs ($R_e = 1.32$ $[1.19-1.47]$, $\frac{1}{\psi} = 7$ $[6-9]$ years), which is in line with underestimation of $R_e$ and $\frac{1}{\psi}$ with bd on the BD-PN simulated data. The likelihood ratio test suggested the BD-PN model: loglikelihood under the BD-PN-estimated parameters was $l_{BDPN} = -1\,140$ vs. $l_{BD}=-1\,146$ under the BD parameters, which gives the ratio of $-2 (l_{BD} - l_{BDPN}) =	12 > 5.99 \approx \chi^2_2(0.95)$. % (the value of chi-squared distribution with 2 degrees of freedom corresponding to the significance level of $0.95$). % unrealistic results: $R_e = 0.5$ (suggesting contained epidemic) and very short infectious time $\frac{1}{\psi} = 0.4$ years. These results were robust across the 10 forests, see \nameref{S1_Table}. As with the results on simulated data, this  highlights the importance of accounting for partner notification during parameter estimation.

@@

In the UK in 2015 the majority of infected population (72\%~\cite{aHIVUnitedKingdom2016}) was on antiretroviral treatment, which, when taken properly, prevents transmission. Hence, the average infectious time roughly represents the average time before starting treatment. Our estimate is approximately 2.3 years.  According to the HIV report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013, less than 3\% of sampled infected individuals were detected though partner notification. Our estimate is even lower, potentially because as our model does not allow for notifying partners during the contact with whom there was no transmission, nor notifying multiple partners.

@@

We applied the PN-test and our estimator to the HIV-1 B MSM cluster in Zurich and to the HIV-1 B epidemic in the UK between 2012 and 2015. The test detected presence of partner notification in both cases. The obtained estimates of epidemiological parameters are in agreement with what we now know about the HIV-1 B epidemic (e.g., $R_e$ between $1.2$ and $1.7$). We estimated a modest partner notification probability ($9\%$ in Zurich, $1\%$ in the UK), and rapid notified partner detection ($\approx 5$ months vs $\approx  8$ years for detection under the standard procedure in Zurich; and $\approx 8$ days vs $\approx 2.3$ years for detection under the standard procedure in the UK). Note that the notification probability represents ``successful notification'', which combines several conditions: the index case remembering and agreeing to notify the partner, the partner being reached, agreeing to take a test, being indeed infected, and moreover the transmission happening during the contact between the index and the partner. This explains the seemingly low values.

@@

The BD-PN model is the first step towards accounting for non-random sampling in phylodynamic models. In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we combined several conditions of the complex partner notification process into one ``successful notification'' probability $\rho$ (see above), and made several additional assumptions: (i) only sampled individuals can notify their partners; (ii) only the most recent partner can get notified; (iii) once notified, partners are always sampled upon removal (however, potentially after the end of the considered sampling period). These assumptions may be too simple to fully grasp the real partner notification process and should be relaxed in the future work. Another needed extension should account for population heterogeneity (i.e., MTBD-PN). As it is much easier to implement a tree simulator with a complex model (e.g., as we did for MTBD-PN models) than to derive its likelihood, a promising path to take for future -PN model developments would be using deep learning (e.g., adding new models to PhyloDeep~\cite{Voznica2021}).

Finally, what we presented here is a specific case of an epidemiological model. However, it might be useful in other contexts as well.
For instance, the State Speciation and Extinction (SSE) family of models~\cite{Maddison2007,fitzjohnEstimatingTraitDependentSpeciation2009,goldbergTempoModePlant2012} in the macroevolutionary domain are mathematically very similar to the epidemiological MTBD models~\cite{macphersonUnifyingPhylogeneticBirth2022}. In macroevolution, a detection of a new species might lead to faster detection of closely related ones. This process could be modeled in the same way as partner notification.

@@

\section*{Acknowledgments}
The authors would like to thank Dr Mathieu Moslonka-Lefebvre who initiated this project by proposing to model partner notification. He worked on a different version of the PN model (which is not part of this manuscript) and on the initial version of the PN test. We would like to thank Dr Jakub Voznica for fruitful discussions on partner notification mechanisms.
We would also like to thank the HPC cluster team in Institut Pasteur for their support with computer simulations.

>>>>

Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}). The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in the maximum-likelihood framework). In particular, it permits the estimation of the following epidemiological parameters:

@@

We propose an %non-Markovian 
extension of the MTBD models that adds partner notification (PN).  
In the partner notification process, an identified infected index patient notifies their potentially infected contacts (e.g., sexual or drug injection partners for HIV), either on their own (patient referral) or with the help of a health service provider (provider referral, when a health adviser notifies the partners without identifying the index patient; or contract referral, when if the index patient fails to notify partners within a specified time period, the health adviser proceeds to provider referral)~\cite{wardPartnerNotification2014}. The notified partners are then provided with appropriate tests and treatment. Therefore, a successful detection of a new case B via partner notification by an index patient A implies several conditions to be met: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (ii) B is successfully reached and notified; (iii) B agrees to take a test; (iv) B is indeed infected. 
For instance, according to the report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013 for HIV, reaching the stage (ii) had an estimated probability of about 50\%; reaching the stage (iii) -- of about 41\% (as 83\% of all the patients who attended a clinic through PN had a reported HIV test); and the stage (iv) -- of less than 3\%.

@@

BD-PN hence has 3 classical BD and 2 additional parameters:
\begin{itemize}
 \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual with this rate. This leads to an internal node in the transmission tree;
 \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This leads to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability. This leads to a tip in the transmission tree being observed;
 \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability. This leads to a removal rate change for the partner branch in the transmission tree;
 \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via the standard procedure) gets removed \textit{and observed} with this rate. This leads to an observed tip in the transmission tree. 
\end{itemize}

@@

We applied bdpn to each of the trees fixing the sampling probability $\rho$ to its true value (for identifiability). %The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material. 
As expected, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings. This shows the consistent behavior of our likelihood calculation and optimization.

@@

The PN test detected partner notification in this tree (p-value $=0.014$). Our bdpn estimator inferred the partner notification probability $\upsilon=0.09$ 95\%-CI $[0.008-0.226]$,  
$R_e = 1.43$ $[1.26-1.66]$, infectious time $\frac{1}{\psi} = 8$ $[6-11]$ years, and 4.9 months [6 days -- 1.4 years] between partner notification and their sampling. These results globally agree (CIs intersect) with the $R_e$ estimate for this data using a coalescent model from Rasmussen \textit{et al.}~\cite{Rasmussen2017} (between 1.0 and 2.5), and the estimates using the BDSS model from Voznica~\textit{et al.}~\cite{Voznica2021}: $R_e$ of 1.6 and 1.7 and an infectious period of 10.2 and 9.8 years (depending on the tree representation used for the deep learner).

@@

For comparison, parameter estimation on the same tree with the BD model gave a slightly lower estimate of $R_e$ and a slightly shorter infectious time, though with intersecting CIs ($R_e = 1.32$ $[1.19-1.47]$, $\frac{1}{\psi} = 7$ $[6-9]$ years), which is in line with underestimation of $R_e$ and $\frac{1}{\psi}$ with the bd estimator on the BD-PN simulated data. The likelihood ratio test suggested the BD-PN model: loglikelihood under the BD-PN-estimated parameters was $l_{BDPN} = -1\,140$ vs. $l_{BD}=-1\,146$ under the BD parameters, which gives the ratio of $-2 (l_{BD} - l_{BDPN}) =	12 > 5.99 \approx \chi^2_2(0.95)$. % (the value of chi-squared distribution with 2 degrees of freedom corresponding to the significance level of $0.95$). % unrealistic results: $R_e = 0.5$ (suggesting contained epidemic) and very short infectious time $\frac{1}{\psi} = 0.4$ years. These results were robust across the 10 forests, see \nameref{S1_Table}. As with the results on simulated data, this  highlights the importance of accounting for partner notification during parameter estimation.

@@

In the UK in 2015 the majority of infected population (72\%~\cite{aHIVUnitedKingdom2016}) was on antiretroviral treatment, which, when taken properly, prevents transmission. Hence, the average infectious time roughly represents the average time before starting treatment. Our estimate is approximately 2.3 years.  According to the HIV report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013, less than 3\% of sampled infected individuals were detected though partner notification. Our estimate is even lower, potentially because our model does not allow for notifying partners with whom there was no transmission, nor notifying multiple partners.

@@

We applied the PN-test and our estimator to the HIV-1 B MSM cluster in Zurich and to the HIV-1 B epidemic in the UK between 2012 and 2015. The test detected presence of partner notification in both cases. The obtained estimates of epidemiological parameters are in agreement with what we now know about the HIV-1 B epidemic (e.g., $R_e$ between $1.2$ and $1.7$). We estimated a modest partner notification probability ($9\%$ in Zurich, $1\%$ in the UK), and rapid notified partner detection ($\approx 5$ months vs $\approx  8$ years for detection under the standard procedure in Zurich; and $\approx 8$ days vs $\approx 2.3$ years for detection under the standard procedure in the UK). Note that the notification probability represents ``successful notification'', which combines several conditions: the index patient remembering and agreeing to notify the partner, the partner being reached, agreeing to take a test, being indeed infected, and moreover the transmission happening during the contact between the index and the partner. This explains the seemingly low values.

@@

The BD-PN model is the first step towards accounting for non-random sampling in phylodynamic models. In order to reduce its mathematical complexity (having closed-form solutions for its differential equations) and make parameter estimation fast, we combined several conditions of the complex partner notification process into one ``successful notification'' probability $\upsilon$ (see above), and made several additional assumptions: (i) only sampled individuals can notify their partners; (ii) only the most recent partner can get notified; (iii) once notified, partners are always sampled upon removal (however, potentially after the end of the considered sampling period). These assumptions may be too simple to fully grasp the real partner notification process and should be relaxed in the future work. Another needed extension should account for population heterogeneity (i.e., MTBD-PN). As it is much easier to implement a tree simulator with a complex model (e.g., as we did for MTBD-PN models) than to derive its likelihood, a promising path to take for future -PN model developments would be using deep learning (e.g., adding new models to PhyloDeep~\cite{Voznica2021}).

Finally, what we presented here is a specific case of an epidemiological model. However, it might be useful in other contexts as well.
For instance, the State Speciation and Extinction (SSE) family of models~\cite{Maddison2007,fitzjohnEstimatingTraitDependentSpeciation2009,goldbergTempoModePlant2012} in the macroevolution domain are mathematically very similar to the epidemiological MTBD models~\cite{macphersonUnifyingPhylogeneticBirth2022}. In macroevolution, a detection of a new species might lead to faster detection of closely related ones. This process could be modeled in the same way as partner notification.

@@

\section*{Acknowledgments}
The authors would like to thank Dr Mathieu Moslonka-Lefebvre who initiated this project by proposing to study partner notification. He worked on a different version of the PN model (which is not part of this manuscript) and on the initial version of the PN test. We would like to thank Dr Jakub Voznica for fruitful discussions on partner notification mechanisms.
We would also like to thank the HPC cluster team in Institut Pasteur for their support with computer simulations. O.G. was supported by the Paris Artificial Intelligence Research Institute (PRAIRIE, ANR-19-P3IA-0001).
